**Machine learning identifies unique taxa differentiating proximal and distal

### 2017-06-12 submission/manuscript.Rmd
first draft for editing done

---
title: '**Machine learning identifies unique taxa differentiating proximal and distal
  human colonic microbiota**'
fontsize: 11pt
geometry: margin=1.0in
output:
  html_document: default
  pdf_document:
    includes: null
    keep_tex: yes
csl: mbio.csl
bibliography: references.bib 
---
\vspace{35mm}

@@

Colorectal cancer (CRC) remains a leading cause of death worldwide. Tumors of the proximal (right) and distal (left) colon are morphologically and genetically distinct. Previous work from our group found that microbial dysbiosis is associated with the development of colorectal cancer tumors in studies of both mice and humans. Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease. In this study, we extended our observations of the fecal microbiome to analysis of the proximal and distal human colon. We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure. This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption. 16S rRNA gene sequencing was performed on proximal and distal mucosal biopsies, luminal and exit stool for 20 healthy individuals. Diversity analysis of each location revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals. Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the machine-learning algorithm Random Forest to identify key species that distinguish biogeographical sites. Random Forest classification models were built using taxa abundance and sample location revealed distinct populations that were found in each location. Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas and Actinomyces were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925). The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755). The left mucosa was found to have high populations of Finegoldia, Murdochiella and Porphyromonas. Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end. Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon. Taken together, our results have identified distinct bacterial populations distinct of the proximal and distal colon. Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States. CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon. Tumors found in the proximal colon are typically flatter, sessile tumors that form along the inside of the colon wall whereas tumors of the distal colon are bulbous and project into the lumen. These features make cancers of the proximal colon less likely to present with bowel symptoms and thus are associated with an increase in morbidity and mortality (@Yamauchi2012). CRC has recently been described to fall into four molecular subtypes (@Guinney2015). Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions that present with painful symptoms. In contrast, right-sided colon tumors are bulky and project into the lumen, often causing no symptoms until they have developed into carcinomas (@Benedix2010). Because of the absence of symptoms, right-sided tumors, which comprise up to 47% of CRCs, have a significantly higher mortality rate (@Benedix2010). The left and right sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015). Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. Given this varied physiology of the proximal-distal axis of the colon, symbiotic microbes and their metabolites likely vary between sites.

@@

#### Microbial membership and diversity of the proximal and distal colon 
Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation prior to the procedure (Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. The relative abundance of each phyla for each site is shown in Figure 2A. This data combines the samples from all participants in the study. Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research. Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on. Thus there was no clear pattern of changes in microbial diversity along the gut axis.

@@

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that are distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances. The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left. The models performed very well when classifying these samples (AUC = 0.797 and AUC = 0.980, respectively). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

@@

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_. It was the only sequence to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. (@Castellarin2011). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample (Figure 6A). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides (Figure 6B). Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Finally, another commonly cancer-associated species, _Parvimonas micra_ was identified to be most similar to OTU248. Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa (Figure 6C). Thus, potentially oncogenic OTUs can be found in our sample set of 20 healthy individuals.

@@

When examining the relative abundance of the different phyla at each site, there is a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes. This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome. Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (@Hong2011). Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (@Stearns2011). That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (@Albenberg2014). Thus, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.

>>>>

---
title: '**Machine learning identifies unique taxa differentiating proximal and distal
  human colonic microbiota**'
csl: mbio.csl
fontsize: 11pt
output:
  word_document: default
  pdf_document:
    includes: null
    keep_tex: yes
  html_document: default
geometry: margin=1.0in
bibliography: references.bib
---
\vspace{35mm}

@@

Colorectal cancer (CRC) remains a leading cause of death worldwide. Tumors of the proximal (right) and distal (left) colon are morphologically and genetically distinct. Previous work from our group found that microbial dysbiosis is associated with the development of CRC tumors in studies of both mice and humans. Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease. In this study, we extended our observations of the fecal microbiome to analysis of the proximal and distal human colon. We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure. This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption. 16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and exit stool for 20 healthy individuals. Diversity analysis revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals. Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the machine-learning algorithm Random Forest to identify key species that distinguish biogeographical sites. Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location. _Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas_ and _Actinomyces_ were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925). The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755). The left mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_. Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end. Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon. Taken together, our results have identified distinct bacterial populations of the proximal and distal colon. Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States. CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon. CRC has recently been described to fall into four molecular subtypes (@Guinney2015). Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions that present with painful symptoms. In contrast, right-sided colon tumors are bulky and project into the lumen, often causing no symptoms until they have developed into carcinomas (@Benedix2010). Because of the absence of symptoms, right-sided tumors, which comprise up to 47% of CRCs, have a significantly higher mortality rate (@Benedix2010). The left and right sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015). Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. Given this varied physiology of the proximal-distal axis of the colon, symbiotic microbes and their metabolites likely vary between sites.

@@

#### Microbial membership and diversity of the proximal and distal colon 
Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation(Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. The relative abundance of each phyla for each site is shown in Figure 2A. This data combines the samples from all participants in the study. Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research. Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa, and so on. Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.

@@

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances. The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left. The models performed well when classifying these samples (AUC = 0.797 and AUC = 0.980, respectively). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

@@

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subsp animalis_. It was the only sequence to align with a _F. nucleatum_ strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. (@Castellarin2011). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample (Figure 6A). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides (Figure 6B). Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Thus, cancer-associated OTUs could be found in our sample set of 20 healthy individuals.

@@

When examining the relative abundance of the different phyla at each site, there is a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes. This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome. Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (@Hong2011). Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (@Stearns2011). That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (@Albenberg2014). Hence, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.
